---
title: "ShinyComplexSurvey: Analysis of Complex Survey Data"
author: "Stalyn Guerrero Gómez"
date: today
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: false
    code-tools: true
    theme: flatly
    highlight-style: github
    df-print: kable
    fig-width: 8
    fig-height: 5
vignette: >
  %\VignetteIndexEntry{ShinyComplexSurvey: Analysis of Complex Survey Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)
```

## Overview {.unnumbered}

**ShinyComplexSurvey** is an R package that provides a tidy-oriented
framework for the analysis of complex survey data. It supports:

- Reading microdata from multiple file formats (CSV, XLSX, SPSS, Stata, RDS)
- Constructing and diagnosing stratified and clustered sampling designs
- Estimating population means, totals, proportions, ratios, and quantiles
  with correct variance estimation under complex designs
- Deriving new variables from existing survey data
- Launching an interactive Shiny application for point-and-click analysis

All estimation functions account for stratification, clustering, and
unequal sampling weights following the linearization approach described in
@lumley2010 and @sarndal1992.

---

## Installation

```{r}
#| label: install
#| eval: false
# Install from GitHub
# install.packages("remotes")
remotes::install_github("stalynGuerrero/ShinyComplexSurvey")
```

```{r}
#| label: load
library(ShinyComplexSurvey)
```

---

## Simulated Data

The package ships with `generate_example_data()`, which generates a
three-level hierarchical dataset (PSUs → households → individuals)
representative of a stratified multistage survey design.

```{r}
#| label: gen-data
data <- generate_example_data(n_upm = 100, seed = 123)
dplyr::glimpse(data)
```

### Variables

| Variable | Type | Description |
|---|---|---|
| `strata` | character | Stratum identifier (S1–S5) |
| `upm` | character | Primary sampling unit (PSU/UPM) |
| `hogar_id` | character | Household identifier |
| `persona_id` | character | Individual identifier |
| `weight` | numeric | Sampling weight ($w_i = 1/\pi_i$) |
| `region` | character | Domain: North / Center / South |
| `sexo` | character | Sex: Male / Female |
| `area` | character | Area: Urban / Rural |
| `edad` | numeric | Age in years |
| `educacion` | factor | Education: Primary / Secondary / Higher |
| `empleo` | factor | Employment: Formal / Informal / Unemployed |
| `ingreso_pc` | numeric | Per-capita household income |
| `gasto_pc` | numeric | Per-capita household expenditure |
| `pobre` | numeric | Poverty indicator (0 / 1) |
| `ingreso2` | numeric | Auxiliary income variable (10 % missing) |

### Hierarchy

The simulation enforces internal consistency:

- Weights are constant within households: $w_{hi} = w_h$
- Per-capita income is constant within households: $y_{hi} = Y_h / N_h$
- Individuals younger than 5 years have `NA` for education

Household income follows a hierarchical gamma model with PSU and
household random effects:

$$
Y_h \sim \text{Gamma}\!\left(\alpha,\; \beta \cdot \exp(u_j + v_h)\right)
$$

where $u_j \sim N(0, 0.09)$ is the PSU effect and $v_h \sim N(0, 0.04)$
is the household effect.

---

## Reading External Data

To load your own microdata use `read_survey_data()`. Supported formats
are detected automatically from the file extension.

```{r}
#| label: read-data
#| eval: false
# CSV
data <- read_survey_data("survey.csv")

# SPSS
data <- read_survey_data("survey.sav")

# Stata
data <- read_survey_data("survey.dta")

# Excel
data <- read_survey_data("survey.xlsx")
```

The function returns the data as a `tibble` and attaches metadata
attributes: `source_path`, `source_format`, `n_rows`, `n_cols`.

---

## Deriving Variables

`mutate_survey_data()` creates new variables from a named list of
one-sided formulas, evaluated sequentially in the environment of the
data frame.

```{r}
#| label: mutate
data <- mutate_survey_data(
  data,
  definitions = list(
    log_ingreso = ~ log(ingreso_pc + 1),
    ratio_gasto = ~ gasto_pc / ingreso_pc
  )
)

dplyr::select(data, ingreso_pc, log_ingreso, ratio_gasto) |> head(4)
```

---

## Survey Design

### Building a Design Object

`as_survey_design_tbl()` wraps `survey::svydesign()` and returns a
`tbl_svy` object compatible with the **srvyr** / **survey** ecosystem.

```{r}
#| label: design
design <- as_survey_design_tbl(
  data    = data,
  weight  = "weight",
  strata  = "strata",
  cluster = "upm",
  nest    = TRUE
)

class(design)
```

Supported configurations:

| Configuration | Arguments |
|---|---|
| Simple random sampling | `weight` only |
| Stratified | `weight` + `strata` |
| Clustered (single-stage) | `weight` + `cluster` |
| Stratified multistage | `weight` + `strata` + `cluster` |
| With FPC | any of the above + `fpc` |

The function validates that weights are strictly positive and free of
missing values, and optionally checks that PSUs are not shared across
strata (`check_psu = TRUE`).

:::{.callout-note}
When a stratum contains only one PSU, variance estimation via Taylor
linearization is undefined. The function automatically sets
`options(survey.lonely.psu = "adjust")` to use the conservative
centered-at-stratum-mean approximation [@cochran1977].
:::

### Diagnosing the Design

```{r}
#| label: describe
describe_survey_design(design)
```

The diagnostic table reports:

| Column | Description |
|---|---|
| `n_obs` | Total sample size |
| `n_strata` | Number of strata |
| `n_clusters` | Number of PSUs |
| `weight_min` | Minimum sampling weight |
| `weight_max` | Maximum sampling weight |
| `weight_mean` | Mean sampling weight |
| `weight_cv` | Coefficient of variation of weights |

---

## Estimation

All estimators are computed with `estimate_survey()`. The function returns
a `tibble` with the following columns:

| Column | Description |
|---|---|
| `variable` | Name of the target variable |
| `estimator` | Type of estimator |
| `estimate` | Point estimate $\hat\theta$ |
| `se` | Standard error $SE(\hat\theta)$ |
| `cv` | Coefficient of variation $CV = SE/\hat\theta$ |
| `deff` | Design effect |
| `lci` | Lower confidence bound |
| `uci` | Upper confidence bound |
| `quality` | Precision label based on CV |

**Precision labels:**

| CV | Label |
|---|---|
| $< 5\%$ | Very high precision |
| $5\%$–$10\%$ | High precision |
| $10\%$–$20\%$ | Acceptable precision |
| $20\%$–$30\%$ | Use with caution |
| $\geq 30\%$ | Low precision |

### Population Mean

The weighted population mean is the Horvitz–Thompson ratio estimator:

$$
\hat{\bar{Y}} = \frac{\sum_{i \in s} w_i y_i}{\sum_{i \in s} w_i}
$$

Variance is estimated by Taylor linearization:

$$
\widehat{Var}(\hat{\bar{Y}}) =
\frac{1}{\hat N^2}\,
\widehat{Var}\!\left(\sum_{i \in s} w_i y_i\right)
$$

```{r}
#| label: mean
r_mean <- estimate_survey(
  design   = design,
  variable = "ingreso_pc",
  estimator = "mean"
)
r_mean
```

### Population Total

The Horvitz–Thompson estimator for the total:

$$
\hat T = \sum_{i \in s} w_i y_i
$$

Under stratified designs:

$$
\hat T = \sum_{h=1}^{H} \sum_{i \in s_h} w_{hi}\, y_{hi}
$$

```{r}
#| label: total
r_total <- estimate_survey(
  design   = design,
  variable = "ingreso_pc",
  estimator = "total"
)
r_total
```

### Proportions

For **binary variables** ($y_i \in \{0, 1\}$), the proportion is
equivalent to the weighted mean:

$$
\hat P = \frac{\sum_{i \in s} w_i y_i}{\sum_{i \in s} w_i}
$$

```{r}
#| label: prop-binary
r_pobre <- estimate_survey(
  design   = design,
  variable = "pobre",
  estimator = "prop"
)
r_pobre
```

For **categorical variables**, the function automatically computes
proportions for each category by constructing indicator variables
$I(y_i = k)$:

```{r}
#| label: prop-cat
r_empleo <- estimate_survey(
  design   = design,
  variable = "empleo",
  estimator = "prop"
)
r_empleo
```

### Ratio Estimator

The ratio estimator compares two population totals:

$$
\hat R = \frac{\sum_{i \in s} w_i y_i}{\sum_{i \in s} w_i x_i}
= \frac{\hat T_Y}{\hat T_X}
$$

Variance is estimated by first-order Taylor linearization [@cochran1977]:

$$
\widehat{Var}(\hat R) \approx
\frac{1}{\hat T_X^2}\,
\widehat{Var}\!\left(\hat T_Y - \hat R\, \hat T_X\right)
$$

**Numeric / Numeric:**

```{r}
#| label: ratio-num
r_ratio <- estimate_survey(
  design      = design,
  estimator   = "ratio",
  numerator   = "ingreso_pc",
  denominator = "gasto_pc"
)
r_ratio
```

**Categorical / Categorical** — e.g. ratio of Formal to Informal workers:

```{r}
#| label: ratio-cat
r_ratio_cat <- estimate_survey(
  design          = design,
  estimator       = "ratio",
  numerator       = "empleo",
  denominator     = "empleo",
  ratio_num_level = "Formal",
  ratio_den_level = "Informal"
)
r_ratio_cat
```

**Numeric / Categorical** — e.g. average income among formal workers:

```{r}
#| label: ratio-mixed
r_ratio_mix <- estimate_survey(
  design          = design,
  estimator       = "ratio",
  numerator       = "ingreso_pc",
  denominator     = "empleo",
  ratio_den_level = "Formal"
)
r_ratio_mix
```

### Quantiles

Quantiles are defined from the weighted empirical cumulative distribution:

$$
\hat F_w(y) = \frac{\sum_{i \in s} w_i\, I(Y_i \le y)}{\sum_{i \in s} w_i}
$$

The $p$-th quantile satisfies $\hat F_w(q_p) = p$. Confidence intervals
are computed using the Woodruff linearization method [@woodruff1952].

```{r}
#| label: quantile
r_quant <- estimate_survey(
  design   = design,
  variable = "ingreso_pc",
  estimator = "quantile",
  probs    = c(0.10, 0.25, 0.50, 0.75, 0.90)
)
r_quant
```

---

## Domain Estimation

Supply the `by` argument to compute estimates within subpopulations.
Variance estimation correctly accounts for the complex design within
each domain (no subsetting bias).

```{r}
#| label: domain-mean
r_region <- estimate_survey(
  design   = design,
  variable = "ingreso_pc",
  estimator = "mean",
  by       = "region"
)
r_region
```

Multiple domain variables are supported:

```{r}
#| label: domain-multi
r_region_area <- estimate_survey(
  design   = design,
  variable = "ingreso_pc",
  estimator = "mean",
  by       = c("region", "area")
)
r_region_area
```

Proportions by domain:

```{r}
#| label: domain-prop
r_pobre_region <- estimate_survey(
  design   = design,
  variable = "pobre",
  estimator = "prop",
  by       = "region"
)
r_pobre_region
```

---

## Formatting Results

`format_results_table()` rounds numeric columns, computes missing CV
or confidence intervals, and converts `conf_level` to a percentage string.

```{r}
#| label: format
format_results_table(r_region, digits = 3)
```

---

## Visualisation

`plot_results_bar()` generates a `ggplot2` bar chart with error bars
representing the confidence interval. Domain variables are detected
automatically as any column not in the standard output columns.

```{r}
#| label: plot-region
#| fig-cap: "Per-capita income by region (mean ± 95% CI)"
plot_results_bar(r_region)
```

```{r}
#| label: plot-multi
#| fig-cap: "Per-capita income by region and area (mean ± 95% CI)"
plot_results_bar(r_region_area)
```

Proportion plots automatically constrain the y-axis to $[0, 1]$:

```{r}
#| label: plot-prop
#| fig-cap: "Poverty rate by region (proportion ± 95% CI)"
plot_results_bar(r_pobre_region)
```

---

## Shiny Application

The package includes a full interactive Shiny application covering the
complete analysis pipeline:

1. **Data** — upload CSV / RDS / XLSX or load built-in example data
2. **Survey design** — choose SRS, stratified, or multistage cluster;
   select weight, stratum, and PSU variables; view MathJax theory
3. **Estimation** — select estimator, target variable, and domains;
   execute and inspect results with precision indicators

Launch the application with:

```{r}
#| label: shiny
#| eval: false
ComplexSurvey_app()
```

---

## Complete Workflow Example

```{r}
#| label: workflow
#| code-fold: true
#| code-summary: "Show complete workflow"

library(ShinyComplexSurvey)

# 1. Generate / load data
data <- generate_example_data(n_upm = 100, seed = 2024)

# 2. Derive new variables
data <- mutate_survey_data(
  data,
  definitions = list(
    log_ingreso = ~ log(ingreso_pc + 1)
  )
)

# 3. Build survey design
design <- as_survey_design_tbl(
  data    = data,
  weight  = "weight",
  strata  = "strata",
  cluster = "upm",
  nest    = TRUE
)

# 4. Diagnose
describe_survey_design(design)

# 5. Estimate mean by domain and format
res <- estimate_survey(
  design    = design,
  variable  = "ingreso_pc",
  estimator = "mean",
  by        = c("region", "area")
)

format_results_table(res, digits = 2)
```

---

## Session Information

```{r}
#| label: session
#| code-fold: true
sessionInfo()
```

---

## References {.unnumbered}

::: {#refs}
:::
