---
title: "intro_shinycomplexsurvey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro_shinycomplexsurvey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(printr)
```

```{r setup}
library(ShinyComplexSurvey)
```


## Overview

`ShinyComplexSurvey` provides a tidy and extensible workflow for the analysis
of complex survey data in R.
The package integrates `survey` and `srvyr` to support:

* complex survey designs (weights, strata, clusters),
* derived variables defined via formulas,
* estimation of means, totals and proportions,
* domain and interaction estimates,
* tidy outputs ready for tables and graphics.

This vignette presents the full workflow using a simulated dataset.

---

## Simulating survey data

We start by simulating microdata with a realistic survey structure including
strata, primary sampling units (PSU), and sampling weights.

```{r}
library(ShinyComplexSurvey)
library(dplyr)
library(tibble)

set.seed(123)

n_strata   <- 3
psu_per_s  <- 4
obs_per_ps <- 20

frame <- expand.grid(
estrato = seq_len(n_strata),
upm     = seq_len(psu_per_s),
obs     = seq_len(obs_per_ps)
)

data <- frame %>%
mutate(
w = runif(n(), 0.5, 2),
miembros = sample(1:6, n(), replace = TRUE),
ingreso = round(
rlnorm(
n(),
meanlog = log(400 + estrato * 100),
sdlog = 0.4
),
0
),
sexo   = sample(c("Hombre", "Mujer"), n(), replace = TRUE),
region = sample(c("Norte", "Centro", "Sur"), n(), replace = TRUE),
etnia  = sample(c("Indigena", "Afro", "Ninguna"), n(), replace = TRUE),
dam    = paste0("DAM_", sample(1:5, n(), replace = TRUE))
) %>%
select(
estrato, upm, w,
sexo, region, etnia, dam,
ingreso, miembros
) %>%
as_tibble()

glimpse(data)
```

## Creating derived variables

Derived variables are defined using formulas and evaluated inside `mutate()`
in a transparent and reproducible way.

```{r}
data <- add_survey_variables(
data,
definitions = list(
ingreso_pc = ~ ingreso / miembros,
pobre      = ~ ingreso_pc < 300
)
)

data %>%
select(ingreso, miembros, ingreso_pc, pobre) %>%
head()
```


## Defining the survey design

The survey design is defined by specifying sampling weights, strata,
and primary sampling units.

```{r}
design <- build_survey_design(
data,
weight  = "w",
strata  = "estrato",
cluster = "upm"
)

summary(design)
describe_survey_design(design)
```



## Estimation

### Global estimates

```{r}
estimate_survey(
design,
variable  = "ingreso_pc",
estimator = "mean"
)

estimate_survey(
design,
variable  = "pobre",
estimator = "prop"
)
```



### Domain estimates

Estimates can be produced for any domain using the `by` argument.

```{r}
estimate_survey(
design,
variable  = "ingreso_pc",
estimator = "mean",
by = "sexo"
)

estimate_survey(
design,
variable  = "pobre",
estimator = "prop",
by = "region"
)
```



### Interaction domains

Multiple grouping variables automatically generate interaction domains.
```{r}
res_prop_dam_etnia <- estimate_survey(
design,
variable  = "pobre",
estimator = "prop",
by = c("dam", "etnia")
)

res_prop_dam_etnia %>% data.frame()
```


## Tables

Results are returned in tidy format and can be formatted for reporting.
```{r}
format_results_table(res_prop_dam_etnia, digits = 2) %>% data.frame()
```


## Plots

Bar charts with confidence intervals adapt automatically to domain
and interaction structures.

```{r}
plot_results_bar(res_prop_dam_etnia)
```

## Summary

`ShinyComplexSurvey` provides a coherent and extensible workflow for
complex survey analysis:

* explicit survey design definition,
* transparent creation of derived variables,
* domain and interaction estimates,
* tidy outputs suitable for reporting and visualization.

The package is designed to support both interactive applications
(e.g. Shiny) and reproducible statistical pipelines.



